---
title: "MSDS692 Practicum I - Online Retail Data Set"
author: "Genet Sebehat"
date: "5/10/2020"
output:
  pdf_document: default
  word_document: default 
  html_document: default
---

```{r setup, include=FALSE} 

# Format R Code Automatically
library(formatR)       

knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
# Set so that long lines in R will be wrapped:
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
options(tinytex.verbose = TRUE) 

``` 


```{r create_tryCatch}
x <- tryCatch(stop("Error"), error = function(e) e)
class(x)  
```

# Setup environment - clearing it out and loading any necessary libraries.
```{r setup_environment}
# Removes all objects to prevent results from previous runs being carried over into new runs
rm(list = ls()) 
# Clears the graphic plots window
graphics.off() 
# Clears the console
cat("\014")     
getwd()
setwd("C:/Users/owner/Desktop/Regis University/MSDS692/Week 1")
```

# R library returns the list of loaded (or available) packages.
```{r load_libraries} 
library(dplyr)
library(car)
library(caret)
library(carData)
library(caTools)
library(e1071)
library(gbm)
library(ggplot2)    
library(lattice)
library(lubridate)  
library(plotly)   
library(randomForest)
library(rpart)
library(rpart.plot) 
library(readxl)   
```

# Load data
```{r load_data}
#specify the path to load the Online Retail excel file
OnlineRetail6 <- read_excel("C:/Users/owner/Desktop/Regis University/MSDS692/Week 1/Online Retail.xlsx")

```

# Create str fuction to display the structure of an R object a compact way 
```{r create_str_function}
# Create str fun
str(OnlineRetail6)
```

# Create colSums fuction
```{r check_NAs}
# any NAs in data set?
colSums(is.na(OnlineRetail6))
```

# Create new variable
```{r add_new_variable}
# Exclude negative quantities
OnlineRetail6 <- subset(OnlineRetail6, Quantity > 0 & UnitPrice > 0)

# Adding a new variable for revenue (unit price * quantity) per line:
OnlineRetail6$Revenue  = with(OnlineRetail6, OnlineRetail6$Quantity*OnlineRetail6$UnitPrice)

# Convert number type variable to double
OnlineRetail6$Revenue <- as.double(OnlineRetail6$Revenue)

```

# Check cancelled invoice
```{r}
# any cancelled invoices?
table(grepl("^C", OnlineRetail6$InvoiceNo))
```

# Create spply function takes data frame as input and gives output in vector or matrix
```{r find_missing_values}
#Check missing values for each column
sapply(OnlineRetail6, function(x) sum(is.na(x)))
```

#Data Preprocessing - handling na's
```{r create_preprocess_data}
#Remove missing value
OnlineRetail7 <- na.omit(OnlineRetail6)

#Check missing values for each column
sapply(OnlineRetail7, function(x) sum(is.na(x)))
```

# Create to use only few rows
```{r create_few_rows}
#Create few rows
OnlineRetail7 <- OnlineRetail7[c(2:1000),]
```

# Create partition
```{r create_split_data}
# Create Partition
index = createDataPartition(y=OnlineRetail7$Revenue, p=0.7, list=FALSE)

# Split to train and test
train = OnlineRetail7[index,]
test = OnlineRetail7[-index,]

```

# Create dim function
```{r create_dim_test}
# create dim fun
dim(train)
dim(test)
```

# Create rf model
```{r create_model1}
# fit the model
set.seed(123)
rf_model = train(Revenue ~., data=train, 
               method="rf", 
               prox=TRUE,
               trControl = trainControl(method = "cv"))

rf_model

```

# Create getTree function
```{r Create_getTree_fun}
# 2ND TREE 
gettree <- getTree(rf_model$finalModel, k=2)
head(gettree, n=10)

```

# Create KNN model
```{r create_model2}
# Create KNN model
set.seed(123)
knn_model = train(Revenue ~., data=train, 
               method="knn", 
               prox=TRUE,
               trControl = trainControl(method = "cv"))

knn_model
```

# Create linear regression model
```{r create_model4}
# fit the model-Linear Regression 
set.seed(123)
lr_model = train(Revenue ~., data=train, 
               method="lm", 
               prox=TRUE,
               trControl = trainControl(method = "cv"))

lr_model

```

# Random forests are usually one of the top performing algorithms along with boosting in any prediction contests.

# They are often difficult to interpret because of these multiple trees that we’re fitting but they can be very accurate for a wide range of problems.

# Create prediction
```{r create_predict}
# Create predict model
pred = predict(rf_model, train)

# add a column in test set for prediction TRUE or FALSE:
train$predRight = pred == train$Revenue
 
# Create table
ptable <- table(pred, train$Revenue)

```

# Create attributes function returns the object's attribute list
```{r create_attributes}
# Create attributes function
attributes(rf_model)
```


